---
title: "02-redes"
author: "George G. Vega Yon"
date: "December 9, 2018"
output: html_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exponential Random Graph Models (ERGMs)

*   ¿Qué conjunto de sub-estructuras dan origen al grafo observado?

*   Por ejemplo: \# de triadas, \# de diadas homofílicas, \# de k-estrellas, etc.

TABLA VA ACA


*   En este caso, $\mathbf{Y} = \mathbf{G}$

*   En general, la distribución de $\mathbf{Y}$ puede ser paremetrizada como sigue:
    
    $$
    \Pr\left(\mathbf{Y}=\mathbf{y}|\theta, \mathcal{Y}\right) = \frac{\exp{\theta^{\mbox{T}}\mathbf{g}(\mathbf{y})}}{\kappa\left(\theta, \mathcal{Y}\right)},\quad\mathbf{y}\in\mathcal{Y}
    \tag{1}
    $$
    
    Donde $\theta\in\Omega\subset\mathbb{R}^q$ es el vactor de coeficientes, y 
    $\mathbf{g}(\mathbf{y})$ es un *q*-vector de estadísticos basados en el
    grafo $\mathbf{y}$.

---

*   El modelo (1) puede ser extendido para incluir covariables (*features*) 
    a nivel nodal o de lazo reemplazando $\mathbf{g}(\mathbf{y})$ por
    $\mathbf{g}(\mathbf{y}, \mathbf{X})$ 
    
*   El denominador,
    
    
    $$
    \kappa\left(\theta,\mathcal{Y}\right) = \sum_{\mathbf{z}\in\mathcal{Y}}\exp{\theta^{\mbox{T}}\mathbf{g}(\mathbf{z})}
    $$
    0
    
    Es la constante de normalización que asegura que (1) es una funcion de probabilidad.

*   Aun restringiendo el conjunto $\mathcal{Y}$ a las redes de tamaño $n$,
    el numero de elementos de $\mathcal{Y}$ hace la estimación de estos modelos
    difícil [@Hunter2008].
    
    $$
    N = 2^{n(n-1)}
    $$
    
## ERGMs: Principales supuestos

*   Modelo generador de datos

*   Dependencia de diadas

*   Red completa

*   Lado práctico: Red relativemente mediana (par de miles de nodos, no más que eso)


## Separable Exponential Random Graph Models (a.k.a. TERGMs)

-   A discrete time model.

-   Estimates a set of parameters $\theta = \{\theta^-, \theta^+\}$ that capture the transition dynamics from $\mathbf{Y}^{t-1}$ to $\mathbf{Y}^{t}$.

-   Assuming that $(\mathbf{Y}^+\perp\mathbf{Y}^-) | \mathbf{Y}^{t-1}$ (the model **dynamic model** is separable), we estimate two models:
    $$
    \begin{align}
    \Pr\left(\mathbf{Y}^+ = \mathbf{y}^+|\mathbf{Y}^{t-1} = \mathbf{y}^{t-1};\theta^+\right),\quad \mathbf{y}^+\in\mathcal{Y}^+(\mathbf{y}^{t-1})\\
    \Pr\left(\mathbf{Y}^- = \mathbf{y}^-|\mathbf{Y}^{t-1} = \mathbf{y}^{t-1};\theta^-\right),\quad \mathbf{y}^-\in\mathcal{Y}^-(\mathbf{y}^{t-1})
    \end{align}
    $$

-   So we end up estimating two ERGMs.
    
## Latent Network Models

-   Social networks are a function of a latent space (literally, xyz for example) $\mathbf{Z}$.

-   Individuals who are closer to each other within $\mathbf{Z}$ have a higher
chance of been connected.

-   Besides of estimating the typical set of parameters $\theta$, a key part of this model is find $\mathbf{Z}$.

-   Similar to TERGMs, under the conditional independence assumption we can estimate:

$$
\Pr\left(\mathbf{Y} =\mathbf{y}|\mathbf{X} = \mathbf{x}, \mathbf{Z}, \theta\right) = \prod_{i\neq j}\Pr\left(y_{ij}|z_i, z_j, x_{ij},\theta\right)
$$

See @hoff2002



## ERGMs: Estimación

*   En el caso general, la estimacion de esta familia de modelos no es factible
    utilizando enumeración exahustiva de los grafos.
    
*   Por ejemplo, si $n = 7$, entonces $\left|\mathcal{Y}\right| = 2^{7(7-1)} \approx 4.39\times10^{12}$

*   Varios métodos propuestos para resolver este problema: Algoritmo Robbins-Monro
    (*stochastic approximation*), MCMC-MLE

*   Revisaremos el último

---

### MCMC-MLE

1.  Initialize the algorithm with an initial guess of $\theta$, call it $\theta^{(t)}$

2.  While (no convergence) do:
    
    a.  Using $\theta^{(t)}$, simulate $M$ networks by means of small changes in the $\mathbf{Y}_{obs}$ (the observed network). This part is done by using an importance-sampling method which weights each proposed network by it's likelihood conditional on $\theta^{(t)}$
    
    b.  With the networks simulated, we can do the Newton step to update the parameter $\theta^{(t)}$ (this is the iteration part in the `ergm` package): $\theta^{(t)}\to\theta^{(t+1)}$
    
    c.  If convergence has been reach (which usually means that $\theta^{(t)}$ and $\theta^{(t + 1)}$ are not very different), then stop, otherwise, go to step a.

For more details see @lusher2012; @admiraal2006; @Snijders2002; @Wang2009 provides details on the algorithm used by PNet (which is the same as the one used in `RSiena`). @lusher2012 provides a short discussion on differences between `ergm` and `PNet`. 


## Context  (cont'd)

\begin{figure}
\centering
\includegraphics[width = .65\linewidth]{plot-graph-4-1.pdf}
\end{figure}

\pause

How can we go beyond descriptive statistics? 

## Small networks and Exponential Random Graph Models

When trying to estimate ERGMs in little networks \pause

*   MCMC fails to converge when trying to estimate a block diagonal (structural
    zeros) model,\pause

*   Same happens when trying to estimate an ERGM for a single (little) graph, \pause

*   Even if it converges, the Asymptotic properties of MLEs are no longer valid
    since the sample size is not large enough.

    
## Rethinking the problem

\pause

*   1st Step: Forget about MCMC-MLE estimation, take advantage of small
    sample and use exact statistic for MLEs: \pause
    
    $$
    \Pr\left(\mathbf{Y}=\mathbf{y}|\theta, \mathcal{Y}\right) = \frac{\exp{\theta^{\mbox{T}}\mathbf{g}(\mathbf{y})}}{\kappa\left(\theta, \mathcal{Y}\right)},\quad\mathbf{y}\in\mathcal{Y}
    $$
    
*   This solves the problem of been able to estimate a small ergm. \pause

*   For this we started working on the `lergm` R package
    (available at https://github.com/USCCANA/lergm):\pause
    
    *   Not from scratch: uses some functions from statnet's `ergm`, in particular
        `ergm-terms`.\pause
    
    *   High performing (up to some point): Some components written in C++\pause
    
    *   Very early stage of development...\pause we'll see if it is worth keep
        working on it!

## Example 1

Let's start by trying to estimate an ERGM for a single graph of size 4

```{r lergm1, echo=TRUE}
library(lergm)
set.seed(12)
x <- sna::rgraph(4)
lergm(x ~ edges + balance + mutual)
```


----------

*   Cool, we are able to estimate ERGMs for little networks! (we actually call
    them lergms), but...\pause
    
*   We still have issues regarding asymptotics.\pause

*   We propose to solve this by using a pooled version of the ERGM


## Solution

*   When estimating a block diagonal ERGM we were essentially assuming
    independence across networks.\pause

*   This means that we can actually do the same with exact statistics approach
    to calculate a joint likelihood:\pause
    
    $$
    \Pr\left(\mathbf{Y}=\{\mathbf{y}_{\color{cyan} i}\}|\theta, \left\{\mathcal{Y}_{\color{cyan}i}\right\}\right) = {\color{cyan} \prod_i} \frac{\exp{\theta^{\mbox{T}}\mathbf{g}(\mathbf{y}_{\color{cyan} i})}}{\kappa_{\color{cyan} i}\left(\theta, \mathcal{Y}_{\color{cyan}i}\right)}
    $$
    \pause
    
*   By estimating a pooled version of the ERGM we can recover the asymptotics 
    of MLEs.\pause
    
*   We implemented this in the `lergm` package
    
## Example 2

Suppose that we have 3 little graphs of sizes 4, 5, and 5:

```{r lergm2, echo=TRUE}
library(lergm)
set.seed(12)
x1 <- sna::rgraph(4)
x2 <- sna::rgraph(5)
x3 <- sna::rgraph(5)

lergm(list(x1, x2, x3) ~ edges + balance + mutual)
```


# Referencias

